Final paper: Towards Proving Termination of \texttt{zelda-mosaic} Algorithms under Conditions on the Input

### 2021-05-02 paper/paper.tex
paper: start theorems

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{array,booktabs}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage[inline]{enumitem}
\newlist{inlist}{enumerate*}{1}
\setlist[inlist]{label= (\alph*)}
\usepackage{mathpartir,semantic}
\usepackage[colorlinks]{hyperref}
\hypersetup{%
    colorlinks=true,
    linkcolor=blue,
    urlcolor=cyan,
    citecolor=blue
}
\usepackage[style=ieee]{biblatex}
\addbibresource{sources.bib}

@@

@@

The important features of the judgements are \(\iname{WF\_Emtpy}\) and
\(\iname{WF\_Cons}\). The judgement \(\iname{WF\_Empty}\) corresponds to the
vacuous case: all elements in an empty list are well-formed with respect to any
shape (there are no such elements!). The judgment \(\iname{WF\_Cons}\) provides
a judgment for increasing the size of the first dimension of a well-formed
matrix by adding a well-formed component to the contents.

@@

\paragraph{Linearly indexing a matrix} Given the operation to linearize a
matrix, one-based linearly indexing is easy. See \S~\ref{S:design_linear_index}
for details on linear indexing. The function \(\func{nth}\) takes a matrix \(m\)
and an index \(n\); it returns the one-based linearly indexed content of \(m\)
at \(n\) (wrapped in the option constructor \(\func{Some}\)) if \(n\) is in
bounds relative to \(m\). When \(n\) is out of bounds, the result is
\(\func{None}\). For one-based indexing with functions that naturally use
zero-based indexing, we use \(n-1\). We should note that the current version of
\(\func{nth}\) has the subtle bug that indexing at \(n=0\) currently returns the
same as indexing at \(n=1\), when it should in fact be out of bounds and thus
\(\func{None}\)---a simple \texttt{if} or \texttt{match} guard would cover this
case.

@@

>>>>

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{array,booktabs}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage[inline]{enumitem}
\newlist{inlist}{enumerate*}{1}
\setlist[inlist]{label= (\alph*)}
\usepackage{mathpartir,semantic}
\usepackage[colorlinks]{hyperref}
\hypersetup{%
    colorlinks=true,
    linkcolor=blue,
    urlcolor=cyan,
    citecolor=blue
}
\usepackage[style=ieee]{biblatex}
\addbibresource{sources.bib}

@@

\theoremstyle{plain} % default
\newtheorem{thm}{Theorem}[section]
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Property}
\newtheorem*{cor}{Corollary}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{example}{Example}[section]
\newtheorem{exercise}{Exercise}[section]
\newtheorem*{prob}{Problem}

\theoremstyle{remark}
\newtheorem*{rem}{Remark}
\newtheorem*{note}{Note}
\newtheorem{case}{Case}

@@

The important features of the judgements are \(\iname{WF\_Empty}\) and
\(\iname{WF\_Cons}\). The judgement \(\iname{WF\_Empty}\) corresponds to the
vacuous case: all elements in an empty list are well-formed with respect to any
shape (there are no such elements!). The judgment \(\iname{WF\_Cons}\) provides
a judgment for increasing the size of the first dimension of a well-formed
matrix by adding a well-formed component to the contents.

@@

\paragraph{Linearly indexing a matrix} Given the operation to linearise a
matrix, one-based linearly indexing is easy. See \S~\ref{S:design_linear_index}
for details on linear indexing. The function \(\func{nth}\) takes a matrix \(m\)
and an index \(n\); it returns the one-based linearly indexed content of \(m\)
at \(n\) (wrapped in the option constructor \(\func{Some}\)) if \(n\) is in
bounds relative to \(m\). When \(n\) is out of bounds, the result is
\(\func{None}\). For one-based indexing with functions that naturally use
zero-based indexing, we use \(n-1\). We should note that the current version of
\(\func{nth}\) has the subtle bug that indexing at \(n=0\) currently returns the
same as indexing at \(n=1\), when it should in fact be out of bounds and thus
\(\func{None}\)---a simple \texttt{if} or \texttt{match} guard would cover this
case.

@@

Unless otherwise stated, the key theorems are predicated on all matrices
involved being well-formed.

The key theorems of our proof development are
\begin{inlist}
\item that the two different definitions of the well-formed property agree
    (Theorem~\ref{Th:well_formed_agree});
\item that \(\func{compute\_shape}\) normalizes by preserving the well-formed
    property (Theorem~\ref{Th:compute_shape_wf_normalizes}); and
\item that \(\func{linearize}\) is correct, \textit{i.e.}, that
    \(\func{linearize'}\) produces a list with size the product of the shape and
    that \(\func{linearize}\) preserves the well-formed property
    (Theorems~\ref{Th:linearize_product} and~\ref{Th:linearize_wf}).
\end{inlist}
We present these, along with a sketch of their proofs, in order.

\begin{thm}[Well-formed definitions agree]\label{Th:well_formed_agree}
    For all matrices \(m\), \(m\) is well-formed under the computational
    definition \(\func{well\_formed}\) if-and-only-if \(m\) is well-formed under
    the inductive definition \(\func{well\_formedI}\). (See
    \texttt{well\_formed\_agree} in the Coq
    development~\cite{zelda_mosaic_proof}.)
\end{thm}

\begin{proof}
    We begin with the forward direction and proceed by induction on the contents
    of \(m\); for each case of the contents, we also consider the cases of the
    shape. Several cases are immediately contradictory, since the resulting
    \(m\) could not have been well-formed in the first place. In the remainder,
    we must prove \(\func{well\_formed}\, m\) implies \(\func{well\_formedI}\,
    m\).

    The base case is easy: simply use \(\iname{WF\_Empty}\).

    In the inductive case, we can use the well-formed hypothesis together with
    the induction hypothesis and auxiliary Lemma~\ref{Lem:wfI_all_wf_t} to
    finish.

    Now the backwards direction. We must prove that \(\func{well\_formedI}\, m\)
    implies \(\func{well\_formed}\, m\). We proceed by induction on the evidence
    for the hypothesis \(\func{well\_formedI}\, m\).

    The base cases are straightforward and follow readily from the definition.

    In the inductive case, we can use the inductive hypotheses to conclude.
\end{proof}

\begin{lem}\label{Lem:wfI_all_wf_t}
    If all elements of the list \(\var{ms}\) are well-formed (using the inductive
    definition) with respect to some shape \(t\), then the following matrix is
    also well-formed: \(\{ \var{shape} = |\var{ms}| :: t; \var{contents} =
    \func{Matrix}\, \var{ms} \}\).
\end{lem}

The proof is straightforward by induction on \(\var{ms}\).

\begin{thm}[Computing the shape normalizes it]\label{Th:compute_shape_wf_normalizes}
    For all matrices \(m\), if \(m\) is well-formed than so is the following
    matrix: \(\{ \var{shape} = \func{compute\_shape}\, m; \var{contents} =
    \func{contents}\, m \}\).
\end{thm}

\begin{proof}
    We proceed by induction on the contents of \(m\); for each case of the
    contents, we also consider the cases of the shape. Several cases are
    immediately contradictory, since the resulting \(m\) could not have been
    well-formed in the first place.

    The base case is straightforward from the definition of
    \(\func{well\_formed}\).

    In the inductive case, we have to deal with a \(\func{Matrix}\) of some list
    \(\var{ms}\). We proceed by case analysis on the list; the empty case is
    easy. When the list is non-empty, we are left to prove the critical
    assumption in the definition of \(\func{compute\_shape}\); namely, that all
    the elements in a well-formed matrix have the same computed shape. This is
    done by application of the induction hypothesis and observing that any
    element in the non-empty \(\var{ms}\) has the same computed shape, thanks to
    Lemma~\ref{Lem:wf_same_shape}.
\end{proof}

\begin{lem}\label{Lem:wf_same_shape}
    Given any two matrices \(m_1, m_2\) that are well-formed with respect to the
    same shape, \(\func{compute\_shape}\, m_1 = \func{compute\_shape}\, m_2\).
\end{lem}

The proof is straightforward by induction on the shape combined with case
analysis on the two matrices.

\begin{thm}[Linearize has resulting shape the product of the original]\label{Th:linearize_product}
\end{thm}

\begin{thm}[Linearize preserves the well-formed property]\label{Th:linearize_wf}
\end{thm}
